### 2022-10-31 ms.Qmd
Initial commit

---
title: dwctaxon, an R package for editing and validating taxonomic data in Darwin Core format
---

## Summary

The dwctaxon R package provides functions to edit and validate taxonomic data in compliance with the widely used Darwin Core (DWC) standard for biological data.
dwctaxon automates several data editing steps, thereby simplifying workflows and decreasing the chance of human-generated data entry errors.
Furthermore, it conducts automated checks to validate taxonomic datasets, thereby alerting database maintainers to problems as soon as possible.
dwctaxon will accelerate the generation and maintenance of taxonomic data, a critical part of the effort to document and conserve earth's biodiversity.

## Statement of need

Taxonomic names, or taxa, are a fundamental unit of biodiversity.
Darwin Core (DWC) is a data standard that specifies how taxa, amongst other biological data, should be formatted.
Darwin Core has been widely adopted by many databases (e.g., GBIF, Catalog of Life, World Flora Online) and provides a common format for databases to interface with each other.

Taxonomic data fundamentally represent scientific hypotheses.
As such, taxonomic data are continuously growing as new species are discovered, old taxonomic concepts are revised, etc.
These changes can be complicated.
For example, changing the status of one name to the synonym of another could potentially involve updating many other names if they are synonyms of the first.
Because of these properties, maintaining a taxonomic database in compliance with DWC can be onerous and error prone if done by hand.
There is therefore a need for software to automate entry and validation of DWC taxonomic data.

The dwctaxon R package addresses this need.
dwctaxon can be used to edit databases by adding new rows or modifying existing data.
dwctaxon automatically updates synonymy, thereby decreasing errors and simplifying workflow.
Furthermore, it runs >15 automated checks to validate that taxonomic data are in compliance with DWC and can be used for taxonomic name resolution.
We anticipate dwctaxon will reduce friction in maintaining taxonomic databases in the face of frequent updates, and be used from smaller projects (e.g., local biodiversity surveys) to large, public databases (e.g., GBIF).

## Features

### Data editing

### Data validation

## Example: GBIF

### 2022-10-31 R/functions.R
Add targets workflow

@@ -1,0 +1,27 @@
+load_backbone <- function(gbif_backbone_zip) {
+
+ # Unzip Taxon.tsv to a temporary directory
+  temp_dir <- tempdir(check = TRUE)
+
+  utils::unzip(
+    gbif_backbone_zip, files = "backbone/Taxon.tsv",
+    overwrite = TRUE, junkpaths = TRUE, exdir = temp_dir)
+
+  # Read in data
+  temp_tsv <- fs::path(temp_dir, "Taxon.tsv")
+  res <- readr::read_tsv(temp_tsv)
+
+  # Delete temp file
+  fs::file_delete(temp_tsv)
+
+  res
+
+}
+
+clean_refs <- function(in_file, out_file) {
+  tempfile <- tempfile()
+  system(glue::glue("sed -e 's/\\\\\\]/\\]/g' {in_file} > {tempfile}"))
+  system(glue::glue("sed -e 's/\\\\\\[/\\[/g' {tempfile} > {out_file}"))
+  fs::file_delete(tempfile)
+  out_file
+}

### 2022-10-31 R/packages.R

@@ -1,0 +1,2 @@
+library(dwctaxon)
+library(tidyverse)

### 2022-10-31 R/process_refs.R

@@ -1,0 +1,26 @@
+# Filter a list of references in YAML format to those cited in an RMD file
+library(tidyverse)
+library(rmdref)
+
+# Filter YAML references from Zotero library.
+# main_library.yaml has been exported from Zotero
+# like this: file -> "export library" -> "Better CSL YAML"
+# or created as a symlink like this:
+# cd _targets/user/data_raw/
+# ln -s ~/Dropbox/bibliography/main_library.yaml
+filter_refs_yaml(
+  rmd_file = "ms.Qmd",
+  yaml_in = "main_library.yaml",
+  yaml_out = "references.yaml", silent = FALSE)
+
+# Write out 'aux' file with citation keys to make collection in Zotero
+# so reference data are easier to work with.
+# The collection can be made with the betterbibtex for Zotero plugin:
+# https://retorque.re/zotero-better-bibtex/citing/aux-scanner/
+extract_citations("ms.Qmd") %>%
+  # aux file just consists of lines formatted like
+  # \citation{KEY}
+  # where KEY is the citation key
+  mutate(latex = glue::glue("\\citation{[key]}", .open = "[", .close = "]")) %>%
+  pull(latex) %>%
+  write_lines("working/dwctaxon_ms.aux")

### 2022-10-31 _targets.R

@@ -1,0 +1,30 @@
+library(targets)
+library(tarchetypes)
+
+source("R/packages.R")
+source("R/functions.R")
+
+tar_plan(
+  tar_file(
+    gbif_backbone_zip_file,
+    "_targets/user/data/backbone.zip"
+  ),
+  gbif_backbone = load_backbone(gbif_backbone_zip_file),
+  gbif_check = dct_validate(
+    gbif_backbone,
+    valid_tax_status = paste(
+      "accepted, doubtful, heterotypic synonym",
+      "homotypic synonym", "proparte synonym",
+      "synonym"),
+    on_fail = "summary"
+  ),
+  tar_quarto(
+    ms_initial,
+    "ms.Qmd"
+  ),
+  # fix reference formatting
+  tar_file(
+    ms_final,
+    clean_refs(ms_initial, "paper.md")
+  )
+)

### 2022-10-31 ms.Qmd
Update MS

@@
 ---
-title: dwctaxon, an R package for editing and validating taxonomic data in Darwin Core format
+format:
+  gfm:
+    wrap: preserve
+editor_options:
+  markdown:
+    wrap: none
+    canonical: true
 ---
+
+```
+title: 'dwctaxon, an R package for editing and validating taxonomic data in Darwin Core format'
+tags:
+  - R
+  - taxonomy
+  - biodiversity
+authors:
+  - name: Joel H. Nitta
+    orcid: 0000-0003-4719-7472
+    affiliation: 1
+  - name: Wataru Iwasaki
+    orcid: 0000-0002-9169-9245
+    affiliation: 1
+affiliations:
+  - name: The University of Tokyo, Japan
+date: 31 October 2022
+bibliography: references.bib
+```
 
@@
 
-The dwctaxon R package provides functions to edit and validate taxonomic data in compliance with the widely used Darwin Core (DWC) standard for biological data.
-dwctaxon automates several data editing steps, thereby simplifying workflows and decreasing the chance of human-generated data entry errors.
+The `dwctaxon` R package provides functions to edit and validate taxonomic data in compliance with the widely used Darwin Core (DWC) standard [@Wieczorek2012] for biological data.
+`dwctaxon` automates several data editing steps, thereby simplifying workflows and decreasing the chance of human-generated data entry errors.
 Furthermore, it conducts automated checks to validate taxonomic datasets, thereby alerting database maintainers to problems as soon as possible.
-dwctaxon will accelerate the generation and maintenance of taxonomic data, a critical part of the effort to document and conserve earth's biodiversity.
+`dwctaxon` will accelerate the generation and maintenance of taxonomic data, a critical part of the effort to document and conserve earth's biodiversity.
 
@@
 
-Taxonomic names, or taxa, are a fundamental unit of biodiversity.
-Darwin Core (DWC) is a data standard that specifies how taxa, amongst other biological data, should be formatted.
-Darwin Core has been widely adopted by many databases (e.g., GBIF, Catalog of Life, World Flora Online) and provides a common format for databases to interface with each other.
+Taxonomic names, or taxa, are a fundamental unit of biodiversity [@Janzen2003].
+Darwin Core (DWC) is a data standard for taxonomic and other biodiversity data, and provides a common format for biodiversity databases to interface with each other [@Wieczorek2012].
+Darwin Core has been widely adopted by many databases (e.g., GBIF, Catalog of Life, World Flora Online).
 
-Taxonomic data fundamentally represent scientific hypotheses.
-As such, taxonomic data are continuously growing as new species are discovered, old taxonomic concepts are revised, etc.
-These changes can be complicated.
-For example, changing the status of one name to the synonym of another could potentially involve updating many other names if they are synonyms of the first.
+Taxonomic data fundamentally represent scientific hypotheses [@Sluys2013].
+As such, taxonomic data are continuously being updated as new species are discovered, old taxonomic concepts are revised, etc.
+Such taxonomic changes can be complicated.
+For example, changing the status of one name to the synonym of another could potentially involve modifying many other names if they are synonyms of the first.
 Because of these properties, maintaining a taxonomic database in compliance with DWC can be onerous and error prone if done by hand.
@@
 
-The dwctaxon R package addresses this need.
-dwctaxon can be used to edit databases by adding new rows or modifying existing data.
-dwctaxon automatically updates synonymy, thereby decreasing errors and simplifying workflow.
+The `dwctaxon` R package addresses this need.
+`dwctaxon` can be used to edit DWC taxonomic databases by adding new rows or modifying existing data in compliance with DWC.
+`dwctaxon` automatically updates synonymy, thereby decreasing errors and simplifying workflow.
 Furthermore, it runs >15 automated checks to validate that taxonomic data are in compliance with DWC and can be used for taxonomic name resolution.
-We anticipate dwctaxon will reduce friction in maintaining taxonomic databases in the face of frequent updates, and be used from smaller projects (e.g., local biodiversity surveys) to large, public databases (e.g., GBIF).
+We anticipate `dwctaxon` will reduce friction in maintaining taxonomic databases in the face of frequent updates, and be used from smaller projects (e.g., local biodiversity surveys) to large, public databases (e.g., GBIF).
 
@@
 
+`dwctaxon` offers several functions for entering new data or manipulating existing data.
+All functions are designed to maximize compliance with DWC.
+The simplest is `dct_add_row()`, which inserts a new row.
+
+`dct_change_status()` modifies the taxonomic status (DWC `taxonomicStatus` term) of an existing row.
+Although DWC does not maintain a standardized list of allowed values for `taxonomicStatus`, these typically include values like "accepted", "synonym", "invalid", etc.
+Changing a name from the status of accepted to synonym can involve several related modifications if the original name included one or more synonyms, as these also need to be updated to point to the new accepted name.
+`dct_change_status()` automatically modifies such synonyms.
+This can reduce error since a given name may include a large number of synonyms that would otherwise be onerous to edit.
+
+`dct_switch_status()` provides a shortcut for a common taxonomic change: swapping accepted/synonym status of two names.
+For example, Species A is a synonym of Species B but in the new taxonomic scheme, we want Species A to be a synonym of Species B.
+`dct_switch_status()` carries out this procedure.
+Like `dct_change_status()`, it automatically re-maps any other synonyms that were formerly associated with the original accepted name.
+
+The default output of all functions is a tibble (data frame).
+This means that a series of changes can be applied using the pipe operator (`|>` or `%>%`).
+Furthermore, the data editing functions can also use a data frame or vector to provide the new data so that entries do not have to be made one per function call.
+
 ### Data validation
 
+`dct_validate()` carries out automated data validation to ensure that the taxonomic database is in compliance with DWC.
+At a minimum, this involves checking that the names of columns (terms) match those used by DWC.
+While DWC makes recommendations for values of some terms, for most of these it does not specify the actual values allowed.
+However, it is typical for taxonomic databases to make assumptions about the data based on their values.
+For example, we would expect that every synonym should be linked to to an accepted name.
+Therefore, validation in `dwctaxon` is flexible, allowing the user to select from multiple checks.
+
+Validation checks in `dwctaxon` include the following (each is a logical argument of `dct_validate()`):
+- `check_taxon_id`: Check that all values of `taxonID` are unique and non-missing.
+- `check_tax_status`: Check that all values of `taxonomicStatus` are valid. Default valid values include "accepted", "synonym", or "variant", but these can be set via the `valid_tax_status` argument.
+- `check_mapping`: Check that all values of `acceptedNameUsageID` map to the `taxonID` of an existing name.
+- `check_mapping_strict`: Check that mapping of names follow expectations based on their taxonomic status (e.g., synonyms must map to accepted names, accepted names cannot have an `acceptedNameUsageID`, etc.).
+- `check_sci_name`: Check that all instances of `scientificName` are non-missing and unique.
+- `check_status_diff`: Check that each scientific name be allowed to have only one taxonomic status (typically only used if `check_sci_name` is not used).
+- `check_col_names`: Check that all column names are valid Darwin Core terms.
+
+The are multiple options for the output of `dct_validate()`:
+- If the `on_fail` argument is set to `error`, an error will be raised and computation stopped at the first failing check.
+- If `on_fail` is set to `summary`, if any check fails a summary table (tibble) will be returning describing failures for all checks, with a warning.
+- If the `on_success` argument is set to `data` and all checks pass, the original data will be returned.
+- If `on_success` is set to `logical` and all checks pass, a logical value of `TRUE` will be returned.
+
+Each check can also be carried out as a separate function (e.g., `dct_check_taxon_id`), with the same output options.
+
 ## Example: GBIF
+
+We demonstrate usage of `dwctaxon` with the GBIF backbone taxonomy.
+
+The taxonomy is loaded into R with `readr::read_tsv()`, then validated with `dct_validate()`.
+```{r}
+#| label: demo-gbif
+#| eval: false
+
+library(dwctaxon)
+
+# This assumes backbone.zip has been downloaded from
+# https://doi.org/10.15468/39omei
+# and unzipped in the current working directory
+
+# Load GBIF backbone taxonomy
+gbif_backbone <- readr::read_tsv("backbone/Taxon.tsv")
+
+# Validate taxonomy
+gbif_check <- dct_validate(
+  gbif_backbone,
+  # - set valid taxonomicStatus as used by GBIF
+  valid_tax_status = paste(
+    "accepted, doubtful, heterotypic synonym,",
+    "homotypic synonym, proparte synonym,",
+    "synonym"),
+  on_fail = "summary")
+```
+
+This reveals a large number of potential issues with the data.
+We see that GBIF uses a non-standard column, `canonicalName`.
+_ scientific names are detected that have duplicates.
+_ names are detected that have an `acceptedNameUsageID` that does not map any `taxonID`.
+It is unknown how these issues currently affect usage of GBIF.
+
+One potential limitation of data validation that relies on in-memory data loaded into R (via `load.csv()` etc.) is that extremely large databases could exceed R's memory capacity.
+However, our demonstration with GBIF data shows that `dwctaxon` functions can be carried out in a reasonable amount of time (in this case, ca.
+20 minutes on an iMac) on a ca.
+2 Gb database that includes a large portion of all scientific names currently in use.
+Therefore, there are likely to be few situations in practice where `dwctaxon` cannot be used due to database size.
+
+## Similar software
+
+The only similar tool we are aware of is the GBIF data validator, an online system for validating data in DWC format.
+However, the GBIF data validator is not available for use as a local program and is limited to input sizes of 100 mb.
+Furthermore, it does not conduct editing of DWC compliant data.
+
+## Conclusion
+
+With its automated yet flexible interface to update and validate taxonomic data in compliance with DWC, `dwctaxon` fills a critical gap in taxonomic data software.
+We hope it will become widely used and support the taxonomic infrastructure that undergirds efforts to document and preserve biodiversity.
+
+## References

### 2022-10-31 R/process_refs.R
Update process_refs.R

@@ -8,13 +8,9 @@ library(rmdref)
 # or created as a symlink like this:
 # cd _targets/user/data_raw/
 # ln -s ~/Dropbox/bibliography/main_library.yaml
-filter_refs_yaml(
-  rmd_file = "ms.Qmd",
-  yaml_in = "main_library.yaml",
-  yaml_out = "references.yaml", silent = FALSE)
 
-# Write out 'aux' file with citation keys to make collection in Zotero
-# so reference data are easier to work with.
+# Write out 'aux' file with citation keys to make collection in Zotero,
+# export to here as bib file.
 # The collection can be made with the betterbibtex for Zotero plugin:
 # https://retorque.re/zotero-better-bibtex/citing/aux-scanner/
 extract_citations("ms.Qmd") %>%

### 2022-11-01 ms.Qmd
Fix yaml header

@@
 
-```
-title: 'dwctaxon, an R package for editing and validating taxonomic data in Darwin Core format'
+\-\-\-
+title: dwctaxon, an R package for editing and validating taxonomic data in Darwin Core format
 tags:
-  - R
-  - taxonomy
-  - biodiversity
+\ \ - R
+\ \ - taxonomy
+\ \ - biodiversity
 authors:
-  - name: Joel H. Nitta
-    orcid: 0000-0003-4719-7472
-    affiliation: 1
-  - name: Wataru Iwasaki
-    orcid: 0000-0002-9169-9245
-    affiliation: 1
+\ \ - name: Joel H. Nitta
+\ \ \ \ orcid: 0000-0003-4719-7472
+\ \ \ \ affiliation: 1
+\ \ - name: Wataru Iwasaki
+\ \ \ \ orcid: 0000-0002-9169-9245
+\ \ \ \ affiliation: 1
 affiliations:
-  - name: The University of Tokyo, Japan
+\ \ - name: The University of Tokyo, Japan
 date: 31 October 2022
 bibliography: references.bib
-```
+\-\-\-
 
@@
 
-The `dwctaxon` R package provides functions to edit and validate taxonomic data in compliance with the widely used Darwin Core (DWC) standard [@Wieczorek2012] for biological data.
+The `dwctaxon` R package provides functions to edit and validate taxonomic data in compliance with the widely used Darwin Core (DWC) standard [@Wieczorek2012] for biodiversity data.
 `dwctaxon` automates several data editing steps, thereby simplifying workflows and decreasing the chance of human-generated data entry errors.
@@
 
-Taxonomic data fundamentally represent scientific hypotheses [@Sluys2013].
+Taxa fundamentally represent scientific hypotheses [@Sluys2013].
 As such, taxonomic data are continuously being updated as new species are discovered, old taxonomic concepts are revised, etc.
@@
 Validation checks in `dwctaxon` include the following (each is a logical argument of `dct_validate()`):
+
 - `check_taxon_id`: Check that all values of `taxonID` are unique and non-missing.
-- `check_tax_status`: Check that all values of `taxonomicStatus` are valid. Default valid values include "accepted", "synonym", or "variant", but these can be set via the `valid_tax_status` argument.
+
+- `check_tax_status`: Check that all values of `taxonomicStatus` are valid. Default valid values include "accepted", "synonym", or 
+"variant", but these can be set via the `valid_tax_status` argument.
+
 - `check_mapping`: Check that all values of `acceptedNameUsageID` map to the `taxonID` of an existing name.
+
 - `check_mapping_strict`: Check that mapping of names follow expectations based on their taxonomic status (e.g., synonyms must map to accepted names, accepted names cannot have an `acceptedNameUsageID`, etc.).
+
 - `check_sci_name`: Check that all instances of `scientificName` are non-missing and unique.
+
 - `check_status_diff`: Check that each scientific name be allowed to have only one taxonomic status (typically only used if `check_sci_name` is not used).
+
 - `check_col_names`: Check that all column names are valid Darwin Core terms.
@@
 The are multiple options for the output of `dct_validate()`:
+
 - If the `on_fail` argument is set to `error`, an error will be raised and computation stopped at the first failing check.
+
 - If `on_fail` is set to `summary`, if any check fails a summary table (tibble) will be returning describing failures for all checks, with a warning.
+
 - If the `on_success` argument is set to `data` and all checks pass, the original data will be returned.
+
 - If `on_success` is set to `logical` and all checks pass, a logical value of `TRUE` will be returned.
@@
 
-## Example: GBIF
+## Example usage
 
@@
 
+## Caveats
+
 One potential limitation of data validation that relies on in-memory data loaded into R (via `load.csv()` etc.) is that extremely large databases could exceed R's memory capacity.
@@
 
-The only similar tool we are aware of is the GBIF data validator, an online system for validating data in DWC format.
+The only similar tool we are aware of is the GBIF data validator (<https://www.gbif.org/tools/data-validator>), an online system for validating data in DWC format.
 However, the GBIF data validator is not available for use as a local program and is limited to input sizes of 100 mb.
-Furthermore, it does not conduct editing of DWC compliant data.
+It does not conduct editing of DWC compliant data, nor are we aware of any other software that does so.
 

### 2022-11-01 R/functions.R

@@ -22,6 +22,7 @@ clean_refs <- function(in_file, out_file) {
   tempfile <- tempfile()
   system(glue::glue("sed -e 's/\\\\\\]/\\]/g' {in_file} > {tempfile}"))
   system(glue::glue("sed -e 's/\\\\\\[/\\[/g' {tempfile} > {out_file}"))
+  system(glue::glue("sed -i '' -e 's/ / /g' {out_file}"))
   fs::file_delete(tempfile)
   out_file
 }

### 2023-12-13 ms.Qmd
Use DwC instead of DWC

@@
 
-The `dwctaxon` R package provides functions to edit and validate taxonomic data in compliance with the widely used Darwin Core (DWC) standard [@Wieczorek2012] for biodiversity data.
+The `dwctaxon` R package provides functions to edit and validate taxonomic data in compliance with the widely used Darwin Core (DwC) standard [@Wieczorek2012] for biodiversity data.
 `dwctaxon` automates several data editing steps, thereby simplifying workflows and decreasing the chance of human-generated data entry errors.
@@
 Taxonomic names, or taxa, are a fundamental unit of biodiversity [@Janzen2003].
-Darwin Core (DWC) is a data standard for taxonomic and other biodiversity data, and provides a common format for biodiversity databases to interface with each other [@Wieczorek2012].
+Darwin Core (DwC) is a data standard for taxonomic and other biodiversity data, and provides a common format for biodiversity databases to interface with each other [@Wieczorek2012].
 Darwin Core has been widely adopted by many databases (e.g., GBIF, Catalog of Life, World Flora Online).
@@
 For example, changing the status of one name to the synonym of another could potentially involve modifying many other names if they are synonyms of the first.
-Because of these properties, maintaining a taxonomic database in compliance with DWC can be onerous and error prone if done by hand.
-There is therefore a need for software to automate entry and validation of DWC taxonomic data.
+Because of these properties, maintaining a taxonomic database in compliance with DwC can be onerous and error prone if done by hand.
+There is therefore a need for software to automate entry and validation of DwC taxonomic data.
 
 The `dwctaxon` R package addresses this need.
-`dwctaxon` can be used to edit DWC taxonomic databases by adding new rows or modifying existing data in compliance with DWC.
+`dwctaxon` can be used to edit DwC taxonomic databases by adding new rows or modifying existing data in compliance with DwC.
 `dwctaxon` automatically updates synonymy, thereby decreasing errors and simplifying workflow.
-Furthermore, it runs >15 automated checks to validate that taxonomic data are in compliance with DWC and can be used for taxonomic name resolution.
+Furthermore, it runs >15 automated checks to validate that taxonomic data are in compliance with DwC and can be used for taxonomic name resolution.
 We anticipate `dwctaxon` will reduce friction in maintaining taxonomic databases in the face of frequent updates, and be used from smaller projects (e.g., local biodiversity surveys) to large, public databases (e.g., GBIF).
@@
 `dwctaxon` offers several functions for entering new data or manipulating existing data.
-All functions are designed to maximize compliance with DWC.
+All functions are designed to maximize compliance with DwC.
 The simplest is `dct_add_row()`, which inserts a new row.
 
-`dct_change_status()` modifies the taxonomic status (DWC `taxonomicStatus` term) of an existing row.
-Although DWC does not maintain a standardized list of allowed values for `taxonomicStatus`, these typically include values like "accepted", "synonym", "invalid", etc.
+`dct_change_status()` modifies the taxonomic status (DwC `taxonomicStatus` term) of an existing row.
+Although DwC does not maintain a standardized list of allowed values for `taxonomicStatus`, these typically include values like "accepted", "synonym", "invalid", etc.
 Changing a name from the status of accepted to synonym can involve several related modifications if the original name included one or more synonyms, as these also need to be updated to point to the new accepted name.
@@
 
-`dct_validate()` carries out automated data validation to ensure that the taxonomic database is in compliance with DWC.
-At a minimum, this involves checking that the names of columns (terms) match those used by DWC.
-While DWC makes recommendations for values of some terms, for most of these it does not specify the actual values allowed.
+`dct_validate()` carries out automated data validation to ensure that the taxonomic database is in compliance with DwC.
+At a minimum, this involves checking that the names of columns (terms) match those used by DwC.
+While DwC makes recommendations for values of some terms, for most of these it does not specify the actual values allowed.
 However, it is typical for taxonomic databases to make assumptions about the data based on their values.
@@
 
-The only similar tool we are aware of is the GBIF data validator (<https://www.gbif.org/tools/data-validator>), an online system for validating data in DWC format.
+The only similar tool we are aware of is the GBIF data validator (<https://www.gbif.org/tools/data-validator>), an online system for validating data in DwC format.
 However, the GBIF data validator is not available for use as a local program and is limited to input sizes of 100 mb.
-It does not conduct editing of DWC compliant data, nor are we aware of any other software that does so.
+It does not conduct editing of DwC compliant data, nor are we aware of any other software that does so.
 
@@
 
-With its automated yet flexible interface to update and validate taxonomic data in compliance with DWC, `dwctaxon` fills a critical gap in taxonomic data software.
+With its automated yet flexible interface to update and validate taxonomic data in compliance with DwC, `dwctaxon` fills a critical gap in taxonomic data software.
 We hope it will become widely used and support the taxonomic infrastructure that undergirds efforts to document and preserve biodiversity.

### 2023-12-15 R/functions.R
Use renv
add function to render JOSS ms

@@ -1,10 +1,10 @@
 load_backbone <- function(gbif_backbone_zip) {
 
- # Unzip Taxon.tsv to a temporary directory
+  # Unzip Taxon.tsv to a temporary directory
   temp_dir <- tempdir(check = TRUE)
 
   utils::unzip(
-    gbif_backbone_zip, files = "backbone/Taxon.tsv",
+    gbif_backbone_zip, files = "Taxon.tsv",
     overwrite = TRUE, junkpaths = TRUE, exdir = temp_dir)
 
   # Read in data
@@ -23,6 +23,31 @@ clean_refs <- function(in_file, out_file) {
   system(glue::glue("sed -e 's/\\\\\\]/\\]/g' {in_file} > {tempfile}"))
   system(glue::glue("sed -e 's/\\\\\\[/\\[/g' {tempfile} > {out_file}"))
   system(glue::glue("sed -i '' -e 's/ / /g' {out_file}"))
+  system(glue::glue("sed -i '' -e 's/“/\"/g' {out_file}"))
+  system(glue::glue("sed -i '' -e 's/”/\"/g' {out_file}"))
   fs::file_delete(tempfile)
   out_file
+}
+
+render_joss_ms <- function(md_file) {
+  # Prepare paths
+  md_file_only <- fs::path_file(md_file)
+  md_file_path <- fs::path(md_file) |>
+    fs::path_abs()
+  md_file_folder <- fs::path_dir(md_file_path)
+  pdf_file_only <- md_file_only |>
+    fs::path_ext_remove() |>
+    fs::path_ext_set("pdf")
+
+  # Run docker
+  babelwhale::run(
+    container_id = "openjournals/inara:latest",
+    volumes = glue::glue("{md_file_folder}:/data"),
+    environment_variables = c("JOURNAL" = "joss"),
+    args = md_file_only,
+    command = "/usr/local/bin/inara"
+  )
+
+  # Return path to output
+  fs::path(md_file_folder, pdf_file_only)
 }

### 2023-12-15 _targets.R

@@ -5,26 +5,41 @@ source("R/packages.R")
 source("R/functions.R")
 
 tar_plan(
-  tar_file(
-    gbif_backbone_zip_file,
-    "_targets/user/data/backbone.zip"
-  ),
-  gbif_backbone = load_backbone(gbif_backbone_zip_file),
+  # Download data ----
+  # - Specify data hash set with contentid
+  # This hash corresponds to the GBIF backbone from 2023-08-28
+  # https://hosted-datasets.gbif.org/datasets/backbone/2023-08-28/backbone.zip
+  gbif_hash = "hash://sha256/2897887cc7fb6fe27188c5b0fa5ce2b14762b43c61ceecd1e3dfad25083f5770", # nolint
+  gbif_backbone = load_backbone(contentid::resolve(gbif_hash, store = TRUE)),
+
+  # Validate data ----
   gbif_check = dct_validate(
     gbif_backbone,
     valid_tax_status = paste(
-      "accepted, doubtful, heterotypic synonym",
-      "homotypic synonym", "proparte synonym",
+      "accepted, doubtful, heterotypic synonym,",
+      "homotypic synonym,", "proparte synonym,",
       "synonym"),
     on_fail = "summary"
   ),
+
+  # Calculate stats ----
+  gbif_backbone_size = nrow(gbif_backbone),
+
+  # Render MS ----
+  # - First render to markdown
   tar_quarto(
-    ms_initial,
-    "ms.Qmd"
+    ms_markdown_initial,
+    "ms.Qmd",
+    quiet = FALSE
   ),
-  # fix reference formatting
+  # - Fix reference formatting
   tar_file(
-    ms_final,
-    clean_refs(ms_initial, "paper.md")
+    ms_markdown_fixed,
+    clean_refs(ms_markdown_initial, "paper.md")
+  ),
+  # - Finally render to JOSS PDF
+  tar_file(
+    ms_pdf,
+    render_joss_ms(ms_markdown_fixed)
   )
 )

### 2023-12-15 ms.Qmd
Update MS, include examples

@@
 
+<!-- YAML header for JOSS paper -->
 \-\-\-
@@
 \ \ \ \ orcid: 0000-0002-9169-9245
-\ \ \ \ affiliation: 1
+\ \ \ \ affiliation: "2, 3, 4"
 affiliations:
-\ \ - name: The University of Tokyo, Japan
-date: 31 October 2022
+\ \ - name: Graduate School of Global and Transdisciplinary Studies, Chiba University, Japan
+\ \ \ \ index: 1
+\ \ - name: Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan
+\ \ \ \ index: 2
+\ \ - name: Department of Biological Sciences, Graduate School of Science, The University of Tokyo, Tokyo, Japan
+\ \ \ \ index: 3
+\ \ - name: Atmosphere and Ocean Research Institute, The University of Tokyo, Chiba, Japan
+\ \ \ \ index: 4
+date: 15 December 2023
 bibliography: references.bib
 \-\-\-
+
+```{r}
+#| label: load
+#| echo: false
+#| include: false
+# Load packages and data from targets pipeline
+suppressPackageStartupMessages({
+  library(targets)
+  library(tidyverse)
+  library(scales)
+  })
+tar_load(gbif_backbone_size)
+tar_load(gbif_check)
+```
 
@@
 `dwctaxon` will accelerate the generation and maintenance of taxonomic data, a critical part of the effort to document and conserve earth's biodiversity.
+`dwctaxon` has passed [code review](https://github.com/ropensci/software-review/issues/574) at [ROpenSci](https://ropensci.org/).
+Complete documentation is available at <https://docs.ropensci.org/dwctaxon/>.
 
@@
 Darwin Core (DwC) is a data standard for taxonomic and other biodiversity data, and provides a common format for biodiversity databases to interface with each other [@Wieczorek2012].
-Darwin Core has been widely adopted by many databases (e.g., GBIF, Catalog of Life, World Flora Online).
+Darwin Core has been widely adopted by many databases (e.g., the [Global Biodiversity Information Facility (GBIF)](https://www.gbif.org), [Catalog of Life](https://www.catalogueoflife.org), and [World Flora Online](https://www.worldfloraonline.org)).
 
 Taxa fundamentally represent scientific hypotheses [@Sluys2013].
-As such, taxonomic data are continuously being updated as new species are discovered, old taxonomic concepts are revised, etc.
-Such taxonomic changes can be complicated.
+As such, taxonomic data are continuously being updated due to, for example, discovery of new species or revision of existing taxonomic concepts.
+These taxonomic changes can be complicated.
 For example, changing the status of one name to the synonym of another could potentially involve modifying many other names if they are synonyms of the first.
-Because of these properties, maintaining a taxonomic database in compliance with DwC can be onerous and error prone if done by hand.
+Because of these properties, maintaining a taxonomic database in compliance with DwC can be onerous and error prone if done by hand, especially if the database is large.
 There is therefore a need for software to automate entry and validation of DwC taxonomic data.
@@
 
+## Installation
+
+The stable version can be installed from [the Comprehensive R Archive Network (CRAN)](https://cran.r-project.org/package=dwctaxon):
+
+```{r}
+#| label: install-cran
+#| eval: false
+install.packages("dwctaxon")
+```
+
+The development version can be installed from [r-universe](https://ropensci.r-universe.dev/dwctaxon):
+
+``` r
+options(repos = c(
+  ropensci = "https://ropensci.r-universe.dev/", 
+  CRAN = "https://cran.rstudio.com/"
+))
+install.packages("dwctaxon", dep = TRUE)
+```
+
 ## Features
@@
 All functions are designed to maximize compliance with DwC.
-The simplest is `dct_add_row()`, which inserts a new row.
 
-`dct_change_status()` modifies the taxonomic status (DwC `taxonomicStatus` term) of an existing row.
+`dct_add_row()` is the simplest editing function, and appends a new row to the data.
+It will automatically generate a unique identifier (`taxonID`) for the row if one is not provided by the user, and only allows entering values for official DwC taxon columns (terms).
+
+`dct_modify_row()` modifies the contents of an existing row, allowing the user to assign new values for any DwC terms (such as `scientificName`, `acceptedNameUsage`, `taxonomicStatus`, etc.).
+For a complete list of DwC terms, see the [Darwin Core Quick Reference Guide](https://dwc.tdwg.org/terms/#taxon).
 Although DwC does not maintain a standardized list of allowed values for `taxonomicStatus`, these typically include values like "accepted", "synonym", "invalid", etc.
 Changing a name from the status of accepted to synonym can involve several related modifications if the original name included one or more synonyms, as these also need to be updated to point to the new accepted name.
-`dct_change_status()` automatically modifies such synonyms.
+`dct_modify_row()` automatically modifies such synonyms.
 This can reduce error since a given name may include a large number of synonyms that would otherwise be onerous to edit.
 
-`dct_switch_status()` provides a shortcut for a common taxonomic change: swapping accepted/synonym status of two names.
-For example, Species A is a synonym of Species B but in the new taxonomic scheme, we want Species A to be a synonym of Species B.
-`dct_switch_status()` carries out this procedure.
-Like `dct_change_status()`, it automatically re-maps any other synonyms that were formerly associated with the original accepted name.
+`dct_fill_col()` automatically fills in the values of one column by matching from another pair of columns.
+This is needed because the DwC format includes some columns that have very closely related data, which may be difficult to manually keep up to date (here called "term-termID pairs").
+For example, there are two columns that specify the accepted name for synonyms: `acceptedNameUsage` (the actual accepted scientific name) and `acceptedNameUsageID` (the unique identifier, or `taxonID`, of the accepted name).
+The reason for having both columns is that since `taxonID` is expected to be unique, it is more accurate as a reference value.
+However, due to the formatting of `taxonID` (for example, a typical value may be a series of numbers and letters like `36HMM`), it is difficult for a human to immediately know what scientific name that `taxonID` refers to; hence it is also helpful to have the name itself available alongside.
+`dct_fill_col()` can be used to automatically fill in such term-termID pairs, which is expected to greatly decrease risk of data entry error.
 
@@
 
-- If the `on_fail` argument is set to `error`, an error will be raised and computation stopped at the first failing check.
+- If the `on_fail` argument is set to `"error"`, an error will be raised and computation stopped at the first failing check.
 
-- If `on_fail` is set to `summary`, if any check fails a summary table (tibble) will be returning describing failures for all checks, with a warning.
+- If `on_fail` is set to `"summary"`, if any check fails a summary table (tibble) will be returning describing failures for all checks, with a warning.
 
-- If the `on_success` argument is set to `data` and all checks pass, the original data will be returned.
+- If the `on_success` argument is set to `"data"` and all checks pass, the original data will be returned.
 
-- If `on_success` is set to `logical` and all checks pass, a logical value of `TRUE` will be returned.
+- If `on_success` is set to `"logical"` and all checks pass, a logical value of `TRUE` will be returned.
 
@@
 
-We demonstrate usage of `dwctaxon` with the GBIF backbone taxonomy.
+First, we create a small fictional dataset in the DwC format (DwC possibly includes many more columns, but we only use a subset here) for demonstration using the `tibble` package:
+
+```{r}
+#| label: load-pkg-data
+#| message: false
+library(tibble)
+library(dwctaxon)
+
+example_data <- tribble(
+  ~taxonID, ~acceptedNameUsageID, ~taxonomicStatus, ~taxonRank, ~scientificName,
+  "id1", NA, "accepted", "species", "Fooius barus",
+  "id2", "id1", "synonym", "species", "Fooius wizum",
+  "id3", NA, "accepted", "species", "Bazius quxus",
+  "id4", NA, "accepted", "species", "Bazius buzbar"
+)
+```
+
+### Data editing
+
+```{r}
+#| label: dct-options
+#| echo: false
+
+# Store existing settings
+old_settings <- dct_options()
+
+# turn off time stamp
+dct_options(stamp_modified = FALSE)
+```
+
+`dct_add_row()` adds one or more rows.
+If the user does not provide a value for `taxonID`, one is generated from the hash of the scientific name, such that it should be unique (as long as the scientific name is unique).
+
+```{r add-row}
+example_data |>
+  dct_add_row(
+    scientificName = "Bazius qixis",
+    taxonomicStatus = "accepted",
+    taxonRank = "species"
+  )
+```
+
+`dct_modify_row()` modifies a row, automatically re-mapping synonyms if needed.
+Here, we change *Bazius buzbar* to be a synonym of *Bazius qixis* (note that `acceptedNameUsageID` is automatically populated):
+
+```{r modify-row}
+example_data |>
+  dct_modify_row(
+    scientificName = "Bazius buzbar",
+    taxonomicStatus = "synonym",
+    acceptedNameUsage = "Bazius quxus"
+  )
+```
+
+`dct_fill_col()` fills in values for columns that have term-termID pairs.
+Here, we demonstrate filling in `acceptedNameUsage` (the resulting values do not appear in the manuscript due to width restrictions, but can be seen by running the code interactively).
+
+```{r fill-col}
+example_data |>
+  dct_fill_col(
+    fill_to = "acceptedNameUsage",
+    fill_from = "scientificName",
+    match_to = "taxonID",
+    match_from = "acceptedNameUsageID"
+  )
+```
+
+### Data validation
+
+The `example_data` dataset is already well-formatted, so it will pass validation:
+
+```{r validate-pass}
+# Default behavior is to return the original dataset if checks pass
+# For this example, return TRUE instead
+dct_validate(example_data, on_success = "logical")
+```
+
+For demonstration purposes, we create a dataset with duplicated and erroneous entries:
+
+```{r make-dirty-filmies}
+# Duplicate some data
+bad_data <- rbind(head(example_data), head(example_data, 2))
+# Replace some values of `acceptedNameUsageID` with random letters
+bad_data$acceptedNameUsageID[sample(seq_along(bad_data), 3)] <- sample(letters, 3)
+```
+
+By default, `dct_validate()` will stop with an error on the first check that fails:
+
+```{r validate-error, error = TRUE}
+dct_validate(bad_data)
+```
+
+But it may be useful to get an overview of all the checks that failed.
+This can be done by setting `on_fail` to `"summary"`:
+
+```{r validate-summary-show, eval = FALSE, echo = TRUE}
+dct_validate(bad_data, on_fail = "summary")
+```
+
+```{r validate-summary-print, echo = FALSE, message = FALSE}
+dct_validate(bad_data, on_fail = "summary") |>
+  dplyr::mutate(error = stringr::str_trunc(error, 40, "right"))
+```
+
+### Piping
+
+All the functions in dwctaxon take a dataframe as their first argument and return a dataframe by default, so they are "pipe-friendly" and can be chained together:
+
+```{r pipe}
+example_data |>
+  dct_modify_row(
+    taxonID = "id2",
+    taxonomicStatus = "accepted"
+  ) |>
+  dct_add_row(
+    scientificName = "Bazius qixis",
+    taxonomicStatus = "accepted"
+  ) |>
+  dct_validate()
+```
+
+It's often a good idea to include `dct_validate()` at the end of a chain to make sure the modified taxonomic database is still correctly formatted.
+
+### Real-world example
+
+The small examples shown above may not reflect the performance of `dwctaxon` when applied to larger "real-world" datasets.
+Therefore, in this section we demonstrate usage of `dwctaxon` with the GBIF backbone taxonomy, which is to our knowledge the largest single database in DwC format currently available online, with `r number(gbif_backbone_size, big.mark = ",")` rows (names) in the most recent release (2023-08-28) as of writing.
 
@@
   gbif_backbone,
-  # - set valid taxonomicStatus as used by GBIF
+  # - set valid values for taxonomicStatus as used by GBIF
   valid_tax_status = paste(
     "accepted, doubtful, heterotypic synonym,",
-    "homotypic synonym, proparte synonym,",
+    "homotypic synonym,", "proparte synonym,",
     "synonym"),
@@
 
-This reveals a large number of potential issues with the data.
-We see that GBIF uses a non-standard column, `canonicalName`.
-_ scientific names are detected that have duplicates.
-_ names are detected that have an `acceptedNameUsageID` that does not map any `taxonID`.
-It is unknown how these issues currently affect usage of GBIF.
+```{r}
+#| label: demo-gbif-hide
+#| echo: false
+warning(
+  targets::tar_meta(fields = warnings, complete_only = TRUE)$warnings %>%
+  str_replace_all("\\.\\. ", ". ") %>%
+  str_split("\\. ") %>%
+  magrittr::extract2(1) %>%
+  paste0(".") %>%
+  map_chr(~str_wrap(., width = 70)) %>%
+  paste(collapse = "\n")
+)
+```
+
+This reveals a large number of potential issues with the data, as shown by the warnings.
+Furthermore, since all potential problems with the data have been saved to a dataframe by setting `on_fail = "summary"`, we can dig into the output for a more detailed view.
+Here, we demonstrate this using the tidyverse:
+
+```{r}
+#| label: error-count-show
+#| message: false
+library(tidyverse)
+gbif_check %>%
+  count(error)
+```
+
+```{r}
+#| label: error-count-hide
+#| echo: false
+gbif_check_count <-
+  gbif_check %>%
+  count(error)
+
+n_err <- function(match_text) {
+  n <- gbif_check_count %>%
+    filter(str_detect(error, match_text)) %>%
+    pull(n) %>%
+    sum()
+  stopifnot(length(n) == 1)
+  if (n > 10) {
+    return(scales::number(n, big.mark = ","))
+  } else {
+    return(as.character(english::english(n)))
+  }
+}
+```
+
+We see that GBIF uses `r n_err("Invalid column names")` non-standard column, `canonicalName`.
+`r n_err("scientificName detected with duplicated")` scientific names have duplicated values.
+`r n_err("acceptedNameUsageID|originalNameUsageID|parentNameUsageID")` names have a value for `acceptedNameUsageID`, `originalNameUsageID`, or `parentNameUsageID` that does not map to any `taxonID`.
+It should be noted that these results do not necessarily indicate that the GBIF database is deficit in practice.
+It is up to the database maintainer to interpret the results of the validation and decide upon the appropriate course of action, if any.
+
+Another real-world example using the Database of Vascular Plants of Canada [@Desmet2013] is available on the `dwctaxon` [documentation website](https://docs.ropensci.org/dwctaxon/articles/real-data.html).
 
@@
 
+```{r}
+#| label: caveats
+#| echo: false
+
+times <- targets::tar_meta(fields = seconds)
+check_time <-
+  times %>%
+  filter(name == "gbif_check") %>%
+  pull(seconds) %>%
+  round(1) %>%
+  paste("seconds")
+```
+
 One potential limitation of data validation that relies on in-memory data loaded into R (via `load.csv()` etc.) is that extremely large databases could exceed R's memory capacity.
-However, our demonstration with GBIF data shows that `dwctaxon` functions can be carried out in a reasonable amount of time (in this case, ca.
-20 minutes on an iMac) on a ca.
+However, our demonstration with GBIF data shows that `dwctaxon` functions can be carried out in a reasonable amount of time (in this case, `r check_time` on an iMac) on a ca.
 2 Gb database that includes a large portion of all scientific names currently in use.
-Therefore, there are likely to be few situations in practice where `dwctaxon` cannot be used due to database size.
+Therefore, there are likely to be few situations where `dwctaxon` cannot be used due to database size.
 
@@
 
+```{r}
+#| label: dct-options-reset
+#| echo: false
+
+# Reset options to previous
+do.call(dct_options, old_settings)
+```
+
+## Acknowledgements
+
+Members of the Iwasaki Lab (The University of Tokyo) provided helpful comments while the package was in development.
+This research supported in part by Japan Society for the Promotion of Science (Kakenhi) grant no. 16H06279, 22H04925 and 22K15171.
+Thanks to Collin Schwantes and Stephen Formel for reviewing the code and Noam Ross for handling the code review process at ROpenSci.
+
 ## References

### 2023-12-15 ms.Qmd
Rearrange examples

@@
 `dwctaxon` has passed [code review](https://github.com/ropensci/software-review/issues/574) at [ROpenSci](https://ropensci.org/).
-Complete documentation is available at <https://docs.ropensci.org/dwctaxon/>.
+Complete documentation is available at <https://docs.ropensci.org/dwctaxon> and source code is available at <https://github.com/ropensci/dwctaxon>.
 
@@
 
-## Example usage
+## Examples and documentation
 
-First, we create a small fictional dataset in the DwC format (DwC possibly includes many more columns, but we only use a subset here) for demonstration using the `tibble` package:
+Examples of code usage, an overview of the DwC taxon data format, and complete description of all functions are available on the `dwctaxon` website, <https://docs.ropensci.org/dwctaxon>.
+The examples include [editing](https://docs.ropensci.org/dwctaxon/articles/editing.html) and [validating](https://docs.ropensci.org/dwctaxon/articles/validation.html) a small dataset that comes with `dwctaxon` as well as a [real-world example](https://docs.ropensci.org/dwctaxon/articles/real-data.html) using the Database of Vascular Plants of Canada [@Desmet2013].
 
-```{r}
-#| label: load-pkg-data
-#| message: false
-library(tibble)
-library(dwctaxon)
+## Benchmark on real data
 
-example_data <- tribble(
-  ~taxonID, ~acceptedNameUsageID, ~taxonomicStatus, ~taxonRank, ~scientificName,
-  "id1", NA, "accepted", "species", "Fooius barus",
-  "id2", "id1", "synonym", "species", "Fooius wizum",
-  "id3", NA, "accepted", "species", "Bazius quxus",
-  "id4", NA, "accepted", "species", "Bazius buzbar"
-)
-```
+One potential limitation of data validation that relies on in-memory data loaded into R (via `load.csv()` etc.) is that extremely large databases could exceed R's memory capacity.
 
-### Data editing
+To test this, we ran `dct_validate()` on the GBIF backbone taxonomy, which is to our knowledge the largest single database in DwC format currently available online with `r number(gbif_backbone_size, big.mark = ",")` rows (names) in the most recent release (2023-08-28) as of writing.
+The test was carried out on an Apple iMac (2019, 16 GB RAM) with R v`r getRversion()` and `dwctaxon` v`r packageVersion("dwctaxon")`.
+The code and results are reproduced below:
 
-```{r}
-#| label: dct-options
-#| echo: false
-
-# Store existing settings
-old_settings <- dct_options()
-
-# turn off time stamp
-dct_options(stamp_modified = FALSE)
-```
-
-`dct_add_row()` adds one or more rows.
-If the user does not provide a value for `taxonID`, one is generated from the hash of the scientific name, such that it should be unique (as long as the scientific name is unique).
-
-```{r add-row}
-example_data |>
-  dct_add_row(
-    scientificName = "Bazius qixis",
-    taxonomicStatus = "accepted",
-    taxonRank = "species"
-  )
-```
-
-`dct_modify_row()` modifies a row, automatically re-mapping synonyms if needed.
-Here, we change *Bazius buzbar* to be a synonym of *Bazius qixis* (note that `acceptedNameUsageID` is automatically populated):
-
-```{r modify-row}
-example_data |>
-  dct_modify_row(
-    scientificName = "Bazius buzbar",
-    taxonomicStatus = "synonym",
-    acceptedNameUsage = "Bazius quxus"
-  )
-```
-
-`dct_fill_col()` fills in values for columns that have term-termID pairs.
-Here, we demonstrate filling in `acceptedNameUsage` (the resulting values do not appear in the manuscript due to width restrictions, but can be seen by running the code interactively).
-
-```{r fill-col}
-example_data |>
-  dct_fill_col(
-    fill_to = "acceptedNameUsage",
-    fill_from = "scientificName",
-    match_to = "taxonID",
-    match_from = "acceptedNameUsageID"
-  )
-```
-
-### Data validation
-
-The `example_data` dataset is already well-formatted, so it will pass validation:
-
-```{r validate-pass}
-# Default behavior is to return the original dataset if checks pass
-# For this example, return TRUE instead
-dct_validate(example_data, on_success = "logical")
-```
-
-For demonstration purposes, we create a dataset with duplicated and erroneous entries:
-
-```{r make-dirty-filmies}
-# Duplicate some data
-bad_data <- rbind(head(example_data), head(example_data, 2))
-# Replace some values of `acceptedNameUsageID` with random letters
-bad_data$acceptedNameUsageID[sample(seq_along(bad_data), 3)] <- sample(letters, 3)
-```
-
-By default, `dct_validate()` will stop with an error on the first check that fails:
-
-```{r validate-error, error = TRUE}
-dct_validate(bad_data)
-```
-
-But it may be useful to get an overview of all the checks that failed.
-This can be done by setting `on_fail` to `"summary"`:
-
-```{r validate-summary-show, eval = FALSE, echo = TRUE}
-dct_validate(bad_data, on_fail = "summary")
-```
-
-```{r validate-summary-print, echo = FALSE, message = FALSE}
-dct_validate(bad_data, on_fail = "summary") |>
-  dplyr::mutate(error = stringr::str_trunc(error, 40, "right"))
-```
-
-### Piping
-
-All the functions in dwctaxon take a dataframe as their first argument and return a dataframe by default, so they are "pipe-friendly" and can be chained together:
-
-```{r pipe}
-example_data |>
-  dct_modify_row(
-    taxonID = "id2",
-    taxonomicStatus = "accepted"
-  ) |>
-  dct_add_row(
-    scientificName = "Bazius qixis",
-    taxonomicStatus = "accepted"
-  ) |>
-  dct_validate()
-```
-
-It's often a good idea to include `dct_validate()` at the end of a chain to make sure the modified taxonomic database is still correctly formatted.
-
-### Real-world example
-
-The small examples shown above may not reflect the performance of `dwctaxon` when applied to larger "real-world" datasets.
-Therefore, in this section we demonstrate usage of `dwctaxon` with the GBIF backbone taxonomy, which is to our knowledge the largest single database in DwC format currently available online, with `r number(gbif_backbone_size, big.mark = ",")` rows (names) in the most recent release (2023-08-28) as of writing.
-
-The taxonomy is loaded into R with `readr::read_tsv()`, then validated with `dct_validate()`.
 ```{r}
@@
 #| echo: false
+# Reproduce warnings picked up during analysis
 warning(
@@
 
-This reveals a large number of potential issues with the data, as shown by the warnings.
-Furthermore, since all potential problems with the data have been saved to a dataframe by setting `on_fail = "summary"`, we can dig into the output for a more detailed view.
-Here, we demonstrate this using the tidyverse:
+```{r}
+#| label: demo-gbif-table-show
+#| eval: FALSE
+# Count types of errors detected
+table(gbif_check$error)
+```
 
 ```{r}
-#| label: error-count-show
-#| message: false
-library(tidyverse)
-gbif_check %>%
-  count(error)
+#| label: demo-gbif-table-hide
+#| echo: false
+#| eval: TRUE
+# Wrap error table names to fit PDF width
+error_table <- table(gbif_check$error)
+error_table_names <-
+  names(error_table) %>%
+  str_wrap(width = 70)
+for (i in seq_along(error_table)) {
+  cat(error_table_names[[i]])
+  cat("\ \ \n")
+  cat(error_table[[i]])
+  cat("\ \ \n")
+}
 ```
@@
 #| echo: false
+
+# Get running times
+times <- targets::tar_meta(fields = seconds)
+check_time <-
+  times %>%
+  filter(name == "gbif_check") %>%
+  pull(seconds) %>%
+  round(1) %>%
+  paste("seconds")
+
 gbif_check_count <-
@@
 
+# Helper function to print number of errors
 n_err <- function(match_text) {
@@
 }
+
 ```
@@
 
-Another real-world example using the Database of Vascular Plants of Canada [@Desmet2013] is available on the `dwctaxon` [documentation website](https://docs.ropensci.org/dwctaxon/articles/real-data.html).
-
-## Caveats
-
-```{r}
-#| label: caveats
-#| echo: false
-
-times <- targets::tar_meta(fields = seconds)
-check_time <-
-  times %>%
-  filter(name == "gbif_check") %>%
-  pull(seconds) %>%
-  round(1) %>%
-  paste("seconds")
-```
-
-One potential limitation of data validation that relies on in-memory data loaded into R (via `load.csv()` etc.) is that extremely large databases could exceed R's memory capacity.
-However, our demonstration with GBIF data shows that `dwctaxon` functions can be carried out in a reasonable amount of time (in this case, `r check_time` on an iMac) on a ca.
-2 Gb database that includes a large portion of all scientific names currently in use.
+The time to run `dct_validate()` even on this relatively large database was quite short: `r check_time`.
 Therefore, there are likely to be few situations where `dwctaxon` cannot be used due to database size.
@@
 
-```{r}
-#| label: dct-options-reset
-#| echo: false
+## Reproducibility
 
-# Reset options to previous
-do.call(dct_options, old_settings)
-```
+Code to generate this manuscript is available at <https://github.com/joelnitta/dwctaxon_ms>.
 

### 2023-12-15 R/functions.R

@@ -25,6 +25,7 @@ clean_refs <- function(in_file, out_file) {
   system(glue::glue("sed -i '' -e 's/ / /g' {out_file}"))
   system(glue::glue("sed -i '' -e 's/“/\"/g' {out_file}"))
   system(glue::glue("sed -i '' -e 's/”/\"/g' {out_file}"))
+  system(glue::glue("sed -i '' -e 's/’/\'/g' {out_file}"))
   fs::file_delete(tempfile)
   out_file
 }

### 2023-12-15 ms.Qmd
Update MS based on input from WI

@@
 
-Taxa fundamentally represent scientific hypotheses [@Sluys2013].
-As such, taxonomic data are continuously being updated due to, for example, discovery of new species or revision of existing taxonomic concepts.
+Taxonomic data are frequently being updated due to, for example, discovery of new species or revision of existing taxonomic concepts.
 These taxonomic changes can be complicated.
 For example, changing the status of one name to the synonym of another could potentially involve modifying many other names if they are synonyms of the first.
-Because of these properties, maintaining a taxonomic database in compliance with DwC can be onerous and error prone if done by hand, especially if the database is large.
+Because of these properties, maintaining a taxonomic database in compliance with DwC can be onerous and error-prone if done by hand, especially if the database is large.
 There is therefore a need for software to automate entry and validation of DwC taxonomic data.
@@
 However, it is typical for taxonomic databases to make assumptions about the data based on their values.
-For example, we would expect that every synonym should be linked to to an accepted name.
+For example, we would expect that every synonym should be linked to an accepted name.
 Therefore, validation in `dwctaxon` is flexible, allowing the user to select from multiple checks.
@@
 
-- `check_mapping_strict`: Check that mapping of names follow expectations based on their taxonomic status (e.g., synonyms must map to accepted names, accepted names cannot have an `acceptedNameUsageID`, etc.).
+- `check_mapping_strict`: Check that mapping of names follows expectations based on their taxonomic status (e.g., synonyms must map to accepted names, accepted names cannot have an `acceptedNameUsageID`, etc.).
 
@@
 
-The are multiple options for the output of `dct_validate()`:
+There are multiple options for the output of `dct_validate()`:
 
@@
 Members of the Iwasaki Lab (The University of Tokyo) provided helpful comments while the package was in development.
-This research supported in part by Japan Society for the Promotion of Science (Kakenhi) grant no. 16H06279, 22H04925 and 22K15171.
+This was research supported in part by Japan Society for the Promotion of Science (Kakenhi) grant no. 16H06279, 22H04925 and 22K15171.
 Thanks to Collin Schwantes and Stephen Formel for reviewing the code and Noam Ross for handling the code review process at ROpenSci.

### 2023-12-15 ms.Qmd
Update MS based on input from WI
Fix rOpenSci capitalization

@@
 `dwctaxon` will accelerate the generation and maintenance of taxonomic data, a critical part of the effort to document and conserve earth's biodiversity.
-`dwctaxon` has passed [code review](https://github.com/ropensci/software-review/issues/574) at [ROpenSci](https://ropensci.org/).
+`dwctaxon` has passed [code review](https://github.com/ropensci/software-review/issues/574) at [rOpenSci](https://ropensci.org/).
 Complete documentation is available at <https://docs.ropensci.org/dwctaxon> and source code is available at <https://github.com/ropensci/dwctaxon>.
@@
 
-Taxonomic data are frequently being updated due to, for example, discovery of new species or revision of existing taxonomic concepts.
+Taxonomic data are frequently updated due to, for example, discovery of new species or revision of existing taxonomic concepts.
 These taxonomic changes can be complicated.
@@
 This was research supported in part by Japan Society for the Promotion of Science (Kakenhi) grant no. 16H06279, 22H04925 and 22K15171.
-Thanks to Collin Schwantes and Stephen Formel for reviewing the code and Noam Ross for handling the code review process at ROpenSci.
+Thanks to Collin Schwantes and Stephen Formel for reviewing the code and Noam Ross for handling the code review process at rOpenSci.
 
